---
title: "part1_q1"
author: "Itzel Espinoza"
date: "2024-11-05"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(dplyr)
library(readr)
library(ggplot2)
library(tidyr)
library(lubridate)
library(knitr)
library(kableExtra)
library(glue)


```


# How COVID-19 case, death rates and Hospitalization rates vary across states

## Summary Table

```{r descriptive_stats, echo=FALSE, include=FALSE}

# --------------
# Loading File
# --------------
state_df <- read_csv("~/Downloads/Covid_project_2024/COVID-19-Outcome-Analysis/covid_mobility_geo_state_daily.csv")

state_df <- as.data.frame(state_df)         # Making it a data frame

state_df <- state_df |>
  mutate(across(-c(statename, stateabbrev), as.double))                 # Making all columns be double type except those that are meant to be character

state_df$date <- as.Date(with(state_df, paste(year, month, day, sep = "-")), 
                                       format = "%Y-%m-%d")

# Check the number of unique states (there are 51 states because it includes District of Columbia)
# print(state_norm_df %>% distinct(statename))

# ----------------------------------
# Filtering df to Pandemic Timeline
# ----------------------------------
# January 20, 2020 first USA COVID-19 outbreak
# May 11, 2023 the pandemic was declared over

pan_state_df <- state_df |>
  filter(date >= as.Date("2020-01-20") & date <= as.Date("2023-05-11"))


# -------------------------------------
# Imputing missing values with averages
# -------------------------------------
# Case rate, death rate, hospitalization rate
pan_state_df  <- pan_state_df|>
  mutate(
    new_case_rate = ifelse(is.na(new_case_rate), mean(new_case_rate, na.rm = TRUE), new_case_rate),
    case_rate = ifelse(is.na(case_rate), mean(case_rate, na.rm = TRUE), case_rate),
    new_death_rate = ifelse(is.na(new_death_rate), mean(new_death_rate, na.rm = TRUE), new_death_rate),
    death_rate = ifelse(is.na(death_rate), mean(death_rate, na.rm = TRUE), death_rate),
    hospitalized_rate = ifelse(is.na(hospitalized_rate), mean(hospitalized_rate, na.rm = TRUE), hospitalized_rate)
  )

# -----------------------------------------------
# Summary statistics by states (mean, median, sd)
# -----------------------------------------------
# COVID-19 case rates, death rates and hospitalization rates
state_sum <- pan_state_df|> 
  group_by(statefips, statename) |> 
  summarize(across(c(new_case_rate, case_rate, new_death_rate, death_rate, hospitalized_rate), 
                   list(mean = ~ mean(.x, na.rm = TRUE), 
                        median = ~ median(.x, na.rm = TRUE), 
                        sd = ~ sd(.x, na.rm = TRUE)))) 

```


```{r, warning=FALSE, echo=FALSE}

# ------------------
# Summary Table 1
# ------------------

state_sum |>
  ungroup() |>  
  select(
    statename,
    case_rate_mean,
    case_rate_median,
    case_rate_sd,
    death_rate_mean,
    death_rate_median,
    death_rate_sd,
    hospitalized_rate_mean,
    hospitalized_rate_median,
    hospitalized_rate_sd
  ) |>
  rename(
    "State Name" = statename,
    "Case Rate Mean" = case_rate_mean,
    "Case Rate Median" = case_rate_median,
    "Case Rate SD" = case_rate_sd,
    "Death Rate Mean" = death_rate_mean,
    "Death Rate Median" = death_rate_median,
    "Death Rate SD" = death_rate_sd,
    "Hospitalized Rate Mean" = hospitalized_rate_mean,
    "Hospitalized Rate Median" = hospitalized_rate_median,
    "Hospitalized Rate SD" = hospitalized_rate_sd
  ) |>
  kbl(
    row.names = TRUE,  # Remove row numbers
    align = "c",
    caption = glue("Table 1: Summary of COVID-19 Case, Death, and Hospitalization Rates Across States"),
    escape = FALSE
  ) |>
  kable_styling(
    bootstrap_options = c("hover", "condensed", "responsive"),
    full_width = TRUE,
    html_font = "Arial"
  )|>
  row_spec(0, color = "white", bold = TRUE, font_size = 14, background = "black") %>%
  column_spec(c(1, 2, 3, 4, 5, 6, 7, 8, 9, 10), border_right = TRUE)

```


## Case rates by state

- Oregon, hawaii,and maine had the lowest case rates of COVID 19
- Rhode Island, north dekota, and alaska had the highest case rates of COVID-19
- The difference between Rhode Island (highest: 21178.01) and Oregon (lowest: 10474.52) was 10703.49 more COVID-19 cases per 100,000 people

```{r plot1, warning=FALSE, echo=FALSE}
# --------------------------------------------------------------
# Plot 1: Mean COVID-19 Case rates by state in ascending order
# --------------------------------------------------------------

ggplot(data = state_sum , aes(x = reorder(statename, case_rate_mean), y = case_rate_mean)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  labs(title = "Mean COVID-19 Case Rate by State",
       x = NULL,
       y = "Mean COVID-19 Case Rate per capita (100,000 people)") +
  theme_classic()+
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

# difference between the state with highest and lowest a.k.a range
# max(state_sum$case_rate_mean) - min(state_sum$case_rate_mean)  

```


## Death rates by state

- Hawaii, Vermont, and Utah had the lowest number of COVID-19 death rates per capita
- New York, Mississippi and new Jersey had the highest number of COVID-19 death rates per capita
- The difference between New Jersey (highest: 287.5914) and Hawaii (lowest: 67.6674) is  219.924 more COVID-19 deaths per 100,000 people

```{r plot2, warning=FALSE, echo=FALSE}
# --------------------------------------------------------------
# Plot 2: Mean COVID-19 death rates by state in ascending order
# --------------------------------------------------------------
# Bar Plot showing Mean COVID-19 Case rates by state in ascending order
ggplot(data = state_sum , aes(x = reorder(statename, death_rate_mean), y = death_rate_mean)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  labs(title = "Mean COVID-19 Death Rate by State",
       x = NULL,
       y = "Mean COVID-19 Death Rate per capita (100,000 people)") +
  theme_classic()+
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

# difference between the state with highest and lowest a.k.a range
# max(state_sum$death_rate_mean) - min(state_sum$death_rate_mean)  

```


## Hospitalization rates by state

- Vermont, Hawaii and Utah had the lowest number of COVID-19 hospitalization rates per capita
- District of Columbia, Delaware, and West Virginia had the highest number of COVID-19 death rates per capita
- The difference between District of Columbia (highest: 26.84725) and Vermont (lowest:  5.644644) is  21.20261 more COVID-19 hospitalizations per 100,000 people

```{r plot3, warning=FALSE, echo=FALSE}
# --------------------------------------------------------------
# Plot 3: Mean COVID-19 hospitalization rates by state in ascending order
# --------------------------------------------------------------
# Bar Plot showing Mean COVID-19 Case rates by state in ascending order
ggplot(data = state_sum , aes(x = reorder(statename, hospitalized_rate_mean), y = hospitalized_rate_mean)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  labs(title = "Mean COVID-19 Hospitalization Rate by State",
       x = NULL,
       y = "Mean COVID-19 Hospitalization Rate per capita (100,000 people)") +
  theme_classic()+
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

# difference between the state with highest and lowest a.k.a range
# max(state_sum$hospitalized_rate_mean) - min(state_sum$hospitalized_rate_mean)  

```


# Covid Case, Deaths and Hospitalization rates relate to Mobility Data

## Average Time Spent Outside of Residential locations by State 

- The District of Columbia, Hawaii, and Maryland exhibited the lowest levels of GPS mobility away from home, suggesting that residents were less likely to travel during the pandemic.
- Montana, Wyoming, and Idaho displayed the highest levels of GPS mobility away from home, indicating that residents were more likely to travel during the pandemic.
- Between District of Columbia (lowest:-3.294372), and Montana (highest:1.761086), there was a difference of 5.055459 standard deviations from each other

```{r, warning=FALSE, echo=FALSE}

# -------------------------------------------------
# Summary statistics by Mobility (mean, median, sd)
# -------------------------------------------------
state_mob_df <- pan_state_df |> 
  group_by(statefips, statename) |> 
  summarize(across(c(gps_away_from_home, gps_retail_and_recreation, gps_transit_stations), 
                   list(mean = ~ mean(.x, na.rm = TRUE), 
                        median = ~ median(.x, na.rm = TRUE), 
                        sd = ~ sd(.x, na.rm = TRUE))))

# ---------------------------------
# Normalizing variables of interest
# ---------------------------------
state_mob_df$gps_away_from_home_mean_norm <- (state_mob_df$gps_away_from_home_mean - mean(state_df$gps_away_from_home, na.rm = TRUE)) / sd(state_mob_df$gps_away_from_home_mean)

state_mob_df$gps_retail_and_recreation_norm <- (state_mob_df$gps_retail_and_recreation_mean - mean(state_df$gps_retail_and_recreation, na.rm = TRUE)) / sd(state_mob_df$gps_retail_and_recreation_mean)

state_mob_df$gps_transit_stations_norm <- (state_mob_df$gps_transit_stations_mean - mean(state_df$gps_transit_stations, na.rm = TRUE)) / sd(state_mob_df$gps_transit_stations_mean)


# --------------------------------------------------------------
# Plot 1: Mean GPS Mobility away from home
# --------------------------------------------------------------
ggplot(data = state_mob_df, aes(x = reorder(statename, gps_away_from_home_mean_norm), 
                                y = gps_away_from_home_mean_norm, 
                                fill = gps_away_from_home_mean_norm > 0)) + 
  geom_bar(stat = "identity") + 
  scale_fill_manual(values = c("TRUE" = "mediumorchid4", "FALSE" = "lightgray")) + 
  labs(title = "Average GPS Mobility Away From Home Across States",
       x = NULL,
       y = "Mean GPS Mobility Away From Home (Normalized)",
       caption = "Bars representing colored light gray indicate states below national average, while those colored purple are above national average") + 
  theme_classic() + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1),
        legend.position = "none")  

# difference between the state with highest and lowest a.k.a range
# max(state_mob_df$gps_away_from_home_mean_norm) - min(state_mob_df$gps_away_from_home_mean_norm)  
```

## Average Time spent at retail and recreation locations by State 

- The District of Columbia, Hawaii, and California exhibited the lowest levels of GPS mobility at retails and recreational areas, suggesting that residents were less likely to travel during the pandemic.
- Alaska, South Dakota and Montana displayed the highest levels of GPS mobility at retails and recreational areas, suggesting that residents were less likely to travel during the pandemic.
- Between District of Columbia (lowest:-3.817296), and Montana (highest: 1.811608), there was a difference of  5.628904 standard deviations from each other
```{r, warning=FALSE, echo=FALSE}
# --------------------------------------------------------------
# Plot 2: Mean GPS Mobility away at retails and recreation areas
# --------------------------------------------------------------
ggplot(data = state_mob_df, aes(x = reorder(statename, gps_retail_and_recreation_norm), 
                                y = gps_retail_and_recreation_norm, 
                                fill = gps_retail_and_recreation_norm > 0)) + 
  geom_bar(stat = "identity") + 
  scale_fill_manual(values = c("TRUE" = "mediumorchid4", "FALSE" = "lightgray")) + 
  labs(title = "Average GPS Mobility at retails and recreation areas Across States",
       x = NULL,
       y = "Mean GPS Mobility at retails and recreation areas (Normalized)",
       caption = "Bars representing colored light gray indicate states below national average, while those colored purple are above national average") + 
  theme_classic() + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1),
        legend.position = "none")  


# difference between the state with highest and lowest a.k.a range
# max(state_mob_df$gps_retail_and_recreation_norm) - min(state_mob_df$gps_retail_and_recreation_norm)  
```

## Average Time at inside transit stations by State 

- The District of Columbia, Massachusetts and Hawaii, exhibited the lowest levels of GPS mobility inside transit stations, suggesting that residents were less likely to travel during the pandemic
- South Dakota, Montana and Wyoming displayed the highest levels of GPS mobility inside transit stations, suggesting that residents were more likely to travel during the pandemic
- Between District of Columbia (lowest:-2.378515), and South Dakota (highest: 1.838154), there was a difference of 4.216669 standard deviations from each other

```{r, warning=FALSE, echo=FALSE}
# --------------------------------------------------
# Plot 3: Mean GPS Mobility inside transit stations
# --------------------------------------------------
ggplot(data = state_mob_df, aes(x = reorder(statename, gps_transit_stations_norm), 
                                y = gps_transit_stations_norm, 
                                fill = gps_transit_stations_norm > 0)) + 
  geom_bar(stat = "identity") + 
  scale_fill_manual(values = c("TRUE" = "mediumorchid4", "FALSE" = "lightgray")) + 
  labs(title = "Average GPS Mobility inside transit stations Across States",
       x = NULL,
       y = "Mean GPS Mobility inside transit stations (Normalized)",
       caption = "Bars representing colored light gray indicate states below national average, while those colored purple are above national average") + 
  theme_classic() + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1),
        legend.position = "none")   


# difference between the state with highest and lowest a.k.a range
# max(state_mob_df$gps_transit_stations_norm) - min(state_mob_df$gps_transit_stations_norm)  


```



# Hospitalization rates across time by state

```{r, warning=FALSE, echo=FALSE}


states_low_deathrate <- state_sum|>
  arrange(death_rate_mean) |>
  head(5) |>
  pull(statename)

states_high_deathrate <- state_sum|>
  arrange(-death_rate_mean) |>
  head(5) |>
  pull(statename)


  

# Sample time series plot
# red color = high death rates 
state_df |> 
  filter(statename %in% c(states_low_deathrate, states_high_deathrate)) |> 
  mutate(line_color = if_else(statename %in% states_high_deathrate, "red", "black")) |>  # Create a color column based on statename
  ggplot(aes(x = date, y = new_death_rate, color = line_color)) +  # Use color column for coloring
  geom_line() + 
  labs(title = "Time Series Plot", x = "Date", y = "Death Rate") + 
  theme_classic() + 
  scale_x_date(
    date_breaks = "1 year",              # Set a tick for each year
    date_labels = "%Y",                  # Show the year as YYYY
    limits = c(min(state_df$date), 
               as.Date(format(max(state_df$date) - 1, "%Y-%m-%d"))), # Adjust the limit to exclude 2025
    breaks = seq(from = min(state_df$date), 
                 to = as.Date(format(max(state_df$date) - 1, "%Y-%m-%d")), 
                 by = "1 year")          # Set breaks for every year within the data range
  ) + 
  facet_wrap(~ statename, scales = "free_y", ncol = 5) +  # Create a separate plot for each statename
  scale_color_identity() +  # Use the color values directly from the 'line_color' column
  theme(axis.text.x = element_text(angle = 90, hjust = 1))



# just loooking at 2

states_low_deathrate <- state_sum|>
  arrange(death_rate_mean) |>
  head(2) |>
  pull(statename)

states_high_deathrate <- state_sum|>
  arrange(-death_rate_mean) |>
  head(2) |>
  pull(statename)




state_df |> 
  filter(statename %in% c(states_low_deathrate, states_high_deathrate)) |>  # Filter for relevant states
  ggplot(aes(x = date, y = new_death_rate, group = statename, color = statename)) +  # Color lines based on statename
  geom_line() + 
  labs(title = "Time Series Plot of Death Rates", x = "Date", y = "Death Rate") + 
  theme_classic() + 
  scale_x_date(
    date_breaks = "1 year",              # Set a tick for each year
    date_labels = "%Y",                  # Show the year as YYYY
    limits = c(min(state_df$date), 
               as.Date(format(max(state_df$date) - 1, "%Y-%m-%d"))), # Adjust the limit to exclude 2025
    breaks = seq(from = min(state_df$date), 
                 to = as.Date(format(max(state_df$date) - 1, "%Y-%m-%d")), 
                 by = "1 year")          # Set breaks for every year within the data range
  ) + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1))  # Rotate x-axis labels for readability


state_df |> 
  filter(statename %in% c(states_low_deathrate, states_high_deathrate)) |>  # Filter for relevant states
  ggplot(aes(x = date, y = new_death_rate, group = statename, color = statename)) +  # Color lines based on statename
  geom_line(stat = "smooth", method = "loess", size = 1) +  # Apply smoothing directly to lines
  labs(title = "Smoothed Time Series Plot of Death Rates", x = "Date", y = "Death Rate") + 
  theme_classic() + 
  scale_x_date(
    date_breaks = "1 year",              # Set a tick for each year
    date_labels = "%Y",                  # Show the year as YYYY
    limits = c(min(state_df$date), 
               as.Date(format(max(state_df$date) - 1, "%Y-%m-%d"))), # Adjust the limit to exclude 2025
    breaks = seq(from = min(state_df$date), 
                 to = as.Date(format(max(state_df$date) - 1, "%Y-%m-%d")), 
                 by = "1 year")          # Set breaks for every year within the data range
  ) + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1))  # Rotate x-axis labels for readability


# Calculate the R-squared value for the linear model
lm_model <- lm(new_case_rate ~ gps_transit_stations, data = state_df %>% filter(statename == 'Georgia', year < 2023, !is.na(gps_away_from_home)))
r_squared <- summary(lm_model)$r.squared

# Create the plot
state_df |> 
  filter(statename == "Georgia", year < 2023, !is.na(gps_away_from_home)) |>  # Filter for Oregon
  ggplot(aes(x = gps_away_from_home, y = new_case_rate)) +  # Set axes
  geom_point() +  # Scatterplot points
  geom_smooth(method = "lm", se = FALSE, color = "blue") +  # Best-fit line without confidence interval
  labs(title = "Scatterplot of Case Rate vs. GPS Away From Home (Oregon)", 
       x = "GPS Away From Home", 
       y = "Case Rate") + 
  theme_minimal() +  # Use a clean, minimal theme
  annotate("text", x = max(state_df$gps_away_from_home) * 0.8, y = max(state_df$new_case_rate) * 0.9, 
           label = paste("R² = ", round(r_squared, 2)), color = "blue", size = 5)  # Add R-squared value as annotation


# Count the number of NA values in the gps_away_from_home column
num_na <- sum(is.na(state_df$gps_away_from_home))

# Print the result
num_na



# Load necessary libraries
library(purrr)

# Function to calculate R-squared for each state
calculate_r_squared <- function(state_name, data) {
  lm_model <- lm(case_rate ~ gps_transit_stations, data = data %>% filter(statename == state_name, year < 2023))
  r_squared <- summary(lm_model)$r.squared
  return(data.frame(state_name = state_name, r_squared = r_squared))
}

# Apply the function to each state and calculate R-squared values
r_squared_values <- state_df %>%
  distinct(statename) %>%  # Get distinct state names
  pull(statename) %>%  # Pull the vector of state names
  map_dfr(~ calculate_r_squared(.x, state_df))  # Apply the function and combine results

# Find the state with the largest R-squared value
state_with_max_r_squared <- r_squared_values %>%
  filter(r_squared == max(r_squared))  # Filter for the state with the highest R-squared

# View the result
state_with_max_r_squared









# Care rate and away from home per day
library(ggplot2)
library(dplyr)
library(tidyr)

# Filter out rows with NA values and select only Oregon
covid_mob_geo_state_filtered <- state_df %>%
  filter(statename == "New Jersey", 
         !is.na(new_case_rate), 
         !is.na(gps_away_from_home), 
         !is.na(date))

# Scale gps_away_from_home to match the range of new_case_rate
min_new_case_rate <- min(covid_mob_geo_state_filtered$new_case_rate, na.rm = TRUE)
max_new_case_rate <- max(covid_mob_geo_state_filtered$new_case_rate, na.rm = TRUE)

covid_mob_geo_state_filtered <- covid_mob_geo_state_filtered %>%
  mutate(
    gps_away_from_home_scaled = (gps_away_from_home - min(gps_away_from_home, na.rm = TRUE)) / 
      (max(gps_away_from_home, na.rm = TRUE) - min(gps_away_from_home, na.rm = TRUE)) *
      (max_new_case_rate - min_new_case_rate) + min_new_case_rate
  )

# Reshape data to long format for plotting
covid_long <- covid_mob_geo_state_filtered %>%
  pivot_longer(cols = c(new_case_rate, gps_away_from_home_scaled), 
               names_to = "variable", 
               values_to = "value")

# Plot the data
ggplot(covid_long, aes(x = date, y = value, color = variable)) +
  geom_point(alpha = 0.7) +                  # Scatter points with slight transparency
  geom_line(alpha = 0.5) +                   # Add a line for better trend visualization
  labs(title = "Scatter Plot of New Case Rate and Scaled GPS Away from Home over Time in Oregon",
       x = "Date",
       y = "Value") +
  theme_classic() +
  scale_color_manual(values = c("new_case_rate" = "blue", "gps_away_from_home_scaled" = "red")) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) # Rotate x-axis labels for readability





```






# Supplemental Plots

## COVID-19 Case, Mortality, and Hospitalization Rates Across 51 States

```{r, warning=FALSE, echo=FALSE}
# ---------------------------------
# Normalizing variables of interest
# ---------------------------------
# normalizing case_rate_mean, death_rate_mean and hospitalization_mean with each other to plot them all together and see how they compare to each other by state  TO DO [make this pplot into a function]

state_sum$case_rate_norm <- (state_sum$case_rate_mean - mean(state_sum$case_rate_mean)) / sd(state_sum$case_rate_mean)

state_sum$death_rate_norm <- (state_sum$death_rate_mean - mean(state_sum$death_rate_mean)) / sd(state_sum$death_rate_mean)

state_sum$hospitalized_rate_norm <- (state_sum$hospitalized_rate_mean - mean(state_sum$hospitalized_rate_mean)) / sd(state_sum$hospitalized_rate_mean)

# --------------------
# Data Transformation
# --------------------
# Reshape the data to long format for ggplot
state_norm_df <- state_sum %>%
  pivot_longer(cols = ends_with("norm"), 
               names_to = "rate_type", 
               values_to = "value")


# Get distinct state names in a sorted order
distinct_state_names <- state_norm_df %>%
  distinct(statename) %>%
  arrange(statename) %>%  # Ensure consistent ordering
  pull(statename)

# Split into the first 25 and the remaining 26
first_25_statename <- head(distinct_state_names, 25)
remaining_26_statename <- tail(distinct_state_names, 26)

# Filter data for the first 25 states
first_25_states <- state_norm_df %>%
  filter(statename %in% first_25_statename)

# Filter data for the remaining 26 states
remaining_26_states <- state_norm_df %>%
  filter(statename %in% remaining_26_statename)

# --------------------
# Plotting Bar Plot
# --------------------
# Function
plot_normalized_rates <- function(data, title_suffix) {
  ggplot(data, aes(x = statename, y = value, fill = rate_type)) +
    geom_bar(stat = "identity", position = "dodge") +
    labs(title = title_suffix,
         x = "State Name",
         y = "Normalized Rate",
         fill = "Rate Type") +
    theme_classic() +
    theme(axis.text.x = element_text(angle = 90, hjust = 1),
          legend.position = "bottom")
}

# Create the plots
plot1 <- plot_normalized_rates(first_25_states, "Covid Case Rates, Death Rates and Hospitalization Rates Across first 25 States")
plot2 <- plot_normalized_rates(remaining_26_states, "Covid Case Rates, Death Rates and Hospitalization Rates Across remaining 25 States")

# Display the plots
plot1
plot2
```